---
title: "Exploring Political and Temporal Influences on WTI Crude Oil Prices"
subtitle: "Bayesian Regression Analysis of the 2024 Pre- and Post-Election Period"
author: 
  - Uma Sadhwani
thanks: "Code and data are available at: [https://github.com/UmaSadhwani/wti-prices](https://github.com/UmaSadhwani/wti-prices)."
date: today
date-format: long
abstract: "This paper analyzes the temporal, political, and economic influences on WTI crude oil prices during the 2024 U.S. Presidential Election. Using Bayesian regression, it explores the relationship between political events, daily market changes, and oil price trends. Results highlight significant effects of the election phase and temporal decline, while short-term fluctuations show limited predictive power."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

The intersection of geopolitics, energy policy, and market dynamics
underscores the critical need to explore the performance of Western
Texas Intermediate (WTI) crude oil prices. The imposition of a proposed
25% tariff on imports from Mexico and Canada, including oil, introduces
a complex ripple effect on trade relationships, inflation, currency
valuations, and broader economic indicators like GDP and interest rates.
While these tariffs aim to exert negotiation leverage, their costs are
unlikely to be absorbed by businesses and will instead be passed on to
end consumers, increasing the price of goods and contributing to
inflationary pressures. For Canada, a key oil exporter, the weakening of
its dollar against the U.S. dollar could amplify these effects, further
straining cross-border trade dynamics and economic cooperation.

The context becomes even more significant when considering the oil
industry’s role in U.S.-Mexico relations. President Trump’s focus on
promoting domestic energy independence through policies like "drill,
baby, drill" and removing regulatory barriers, alongside rhetoric about
the U.S. being the "liquid gold" leader, positions oil as both an
economic tool and a geopolitical instrument. At the same time, sanctions
on Iranian and Venezuelan oil create tighter global markets, potentially
bolstering U.S. exports to Southeast Asia while driving domestic
production.

However, trade wars and tariffs pose risks to oil demand and broader
economic stability. Retaliatory measures by Mexico and Canada could
complicate energy exports, while the auto industry, a major consumer of
petroleum products, may face significant disruptions from increased
material costs. With tariffs not set to take effect until early 2025,
governments and businesses face a window of uncertainty to negotiate or
prepare for potential fallout.

Amidst these dynamics, OPEC’s production cuts and the sustained low oil
prices (hovering around \$72–\$75 per barrel) further complicate the
outlook. Although U.S. producers have hedged against price volatility,
the long-term sustainability of elevated production levels remains
uncertain. This backdrop highlights the need to understand the variables
influencing WTI crude oil prices, including election cycles, tariff
policies, and global market forces, to better navigate the interplay
between energy markets and economic policy.

By exploring the temporal trends, election impacts, and daily
fluctuations in WTI crude oil prices, this research seeks to shed light
on the dynamics underpinning one of the most influential commodities in
the global economy. Such insights are essential for informed
decision-making in a world increasingly shaped by energy market
transformations.

This statistical paper aims to analyze the factors influencing the
performance of Western Texas Intermediate (WTI) crude oil prices, with a
specific focus on the interplay between political events, market
dynamics, and temporal trends. The primary objective is to model the
daily price fluctuations of WTI crude oil using a combination of
temporal, economic, and geopolitical predictors, including the date,
election phase, and measures of short-term price changes. By
incorporating these variables, the paper seeks to uncover patterns and
relationships that offer insights into the drivers of oil price
volatility and broader market behavior.

Western Texas Intermediate (WTI) is one of the three primary benchmarks
for global crude oil pricing, alongside Brent crude and Dubai/Oman. It
is considered the leading price reference for oil extracted and traded
in North America, particularly due to its origin at Cushing, Oklahoma, a
major hub for U.S. oil storage and distribution. WTI is widely regarded
as a critical indicator of supply and demand dynamics not only within
the U.S. but also globally, reflecting shifts in production levels,
consumption patterns, geopolitical tensions, and broader economic
trends.

A significant component of this research is the exploration of the 2024
U.S. Presidential Election as a potential inflection point in WTI
prices. The election phase variable is included to investigate whether
political transitions, particularly under a pro-oil administration,
influence market confidence, production strategies, or regulatory
outlooks. Additionally, the paper examines how short-term measures of
market volatility- daily percentage changes and absolute price changes-
relate to longer-term price movements.

The analysis is conducted using Bayesian regression techniques, offering
a probabilistic framework to assess the uncertainty and significance of
the predictors. The ultimate aim is to provide a robust statistical
foundation for understanding the economic and political factors that
shape crude oil markets, addressing both immediate price fluctuations
and overarching trends. This research contributes to broader discussions
about energy market dynamics, the impact of trade policies (like
tariffs), and the intersection of geopolitical events with economic
outcomes, offering valuable insights for policymakers, investors, and
industry stakeholders.

This report investigates Crude Oil Prices: West Texas Intermediate (WTI)
stock performance from the period of October 1st, 2024 (pre-election) to
November 18th, 2024 (post-election).

Our findings indicate that factors like election phase and date numeric
significantly predict fluctuations in WTI crude oil prices, while daily
price change and absolute price change play a less substantial role. For
the election phase, our analysis reveals a mean effect of -2.8 with a
standard deviation of 1.1, indicating that WTI prices are \$2.8 lower on
average during the pre-election phase compared to the post-election
phase. The consistent negative effect, supported by the confidence
interval, highlights the influence of elections on market pricing.

For date numeric, the mean effect is -0.2 with a standard deviation of
0.0, suggesting that WTI prices decrease by \$0.2 per day on average.
While small, this downward trend is statistically significant, capturing
a gradual decline in prices over time.

In contrast, daily percentage change has a mean of 0.5 and a standard
deviation of 1.5. The small coefficient, with a credible interval
spanning zero, suggests no strong or consistent relationship between
daily percentage changes and overall price levels. Similarly, for
absolute price change, the mean effect is -0.3 with a standard deviation
of 2.0, indicating that its influence on WTI prices is not statistically
significant, as the credible interval includes both positive and
negative values.

These findings collectively emphasize the importance of temporal and
political factors in shaping crude oil prices, while short-term
fluctuations appear to have limited predictive value.

Our analysis reveals several noteworthy patterns in WTI crude oil price
dynamics. Notably, prices are systematically lower during the
pre-election phase compared to the post-election phase, suggesting the
influence of market uncertainties or anticipatory sentiment tied to the
electoral cycle. In contrast, short-term fluctuations, whether measured
as daily percentage changes or absolute price changes, exhibit high
variability but lack a consistent relationship with overall price
levels. The model's robust fit, supported by convergence diagnostics
(Rhat = 1) and strong alignment between observed and predicted values,
underscores the reliability of these findings and provides a solid
foundation for further exploration of the factors shaping crude oil
markets.

Telegraphing paragraph: The remainder of this paper is structured as
follows. @sec-data....

# Data {#sec-data}

## 2.1 Overview

For this analysis, we used the R programming language (R Core Team,
2023) to explore the stock performance of Western Texas Intermediate
(WTI) crude oil, based in Cushing, Oklahoma. The dataset, sourced from
the St. Louis Federal Reserve's FRED database, provides historical daily
price data for WTI crude oil, spanning a defined period. This dataset
offers a rich basis for understanding crude oil price trends, short-term
fluctuations, and broader market behaviors, all of which are crucial in
the context of energy economics. WTI crude oil prices are widely
considered a benchmark in the global energy market and have significant
implications for financial markets, geopolitical stability, and
macroeconomic policy.

The dataset includes key variables such as wti_price, which represents
the daily closing price of WTI crude oil in USD, and date, a timestamp
of each observation. To better capture the dynamics of crude oil price
movements, additional variables were constructed during preprocessing.
These include:

```{r}
# Load necessary libraries
library(tidyverse)
library(knitr)

# File path to the dataset
file_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load the dataset
analysis_data <- read_csv(file_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )

# Compute summary statistics
summary_stats <- analysis_data %>%
  summarise(
    `Daily % Change` = paste0(round(mean(daily_pct_change, na.rm = TRUE), 2), " ± ", 
                              round(sd(daily_pct_change, na.rm = TRUE), 2)),
    `Price Change` = paste0(round(mean(price_change, na.rm = TRUE), 2), " ± ", 
                            round(sd(price_change, na.rm = TRUE), 2)),
    `Date Numeric` = paste0(round(mean(date, na.rm = TRUE), 2), " ± ", 
                            round(sd(date, na.rm = TRUE), 2)),
    `Election Phase (Pre vs. Post)` = paste0(
      sum(analysis_data$election_phase == "Pre-Election", na.rm = TRUE), " Pre | ",
      sum(analysis_data$election_phase == "Post-Election", na.rm = TRUE), " Post"
    )
  )

# Convert to a readable table using knitr
summary_stats %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "Variable") %>%
  rename(`Summary Statistics` = V1) %>%
  kable(caption = "Summary Statistics of Variables")
```

Table 1. Summary Statistics of Key Variables in the Analysis.

This information is significant because it serves as a stand-in for
comprehending how the energy industry affects the overall economy.
Policymakers, investors, and economists must pay close attention to WTI
crude oil prices because they have an impact on inflation, currency
valuations, trade balances, and industrial costs. Since WTI is a product
that is mostly sold in US dollars, changes in its price have
geopolitical ramifications that impact the economies of both countries
that produce and import oil.

The FRED database was used for this analysis because of its
accessibility, extensive historical coverage, and reputation for
dependability, even though equivalent statistics are available from
other sources like Bloomberg or the U.S. Energy Information
Administration (EIA).

Calculating percentage and absolute price changes, classifying
observations by election phase, and transforming the date variable to
numeric form for modeling were examples of high-level pre-processing
tasks. Pre-processing preserved the dataset's integrity and pertinence
to the study objectives while guaranteeing that it was prepared for
Bayesian regression analysis.

The analysis's main focus is wti_price, which represents the Western
Texas Intermediate (WTI) daily closing prices. The date was used to find
systematic changes and infer temporal trends. The metrics price_change
and daily_pct_change are used to measure short-term market volatility.
election_phase was added to investigate how political events could
affect prices.

## 2.2 Measurement

To comprehend how actual economic and geopolitical events are converted
into structured data within the dataset on Western Texas Intermediate
(WTI) crude oil prices, a detailed discussion of measurement is required
for this project. Key elements of the global oil market, including
supply-demand dynamics, geopolitical events, and short-term market
volatility, are reflected in the dataset and are methodically recorded
and categorized.

The collection starts with the WTI crude oil daily closing prices, which
are taken from the FRED database maintained by the St. Louis Federal
Reserve. Real-time trends in crude oil markets are reflected in these
pricing statistics, which are impacted by sanctions, geopolitical
events, global economic conditions, and OPEC output decisions.

To facilitate a structured analysis, several key variables were
constructed within the dataset. The wti_price variable represents the
daily closing price of WTI crude oil in USD, serving as the primary
dependent variable. date_numeric provides a numeric representation of
the date, tracking the number of days since the dataset's inception,
enabling temporal trend analysis. daily_pct_change measures the
percentage change in price relative to the previous day, capturing
day-to-day market volatility, while price_change represents the absolute
dollar change from the previous day, offering an alternative measure of
short-term price shifts. Lastly, election_phase categorizes the data
into "pre-election" and "post-election" periods surrounding the 2024
U.S. Presidential Election, allowing for an examination of political
cycles' effects on crude oil performance.

These variables collectively provide a systematic framework to quantify
crude oil price behaviour and assess their responsiveness to underlying
economic and political drivers. This approach enables a deeper
understanding of the dynamic factors shaping the global energy market.

The dataset was generated with a number of important variables to aid in
a systematic study. The main dependent variable is the wti_price
variable, which is the daily closing price of WTI crude oil in USD.
Temporal trend analysis is made possible by date_numeric, which gives a
numeric representation of the date and counts the days since the
dataset's creation. While price_change shows the absolute dollar change
from the previous day, providing an alternate measure of short-term
price movements, daily_pct_change measures the percentage change in
price relative to the previous day, capturing day-to-day market
volatility. In order to examine the impact of political cycles on crude
oil performance, election_phase lastly divides the data into
"pre-election" and "post-election" phases surrounding the 2024 U.S.
Presidential Election.

Together, these factors offer a methodical framework for measuring the
evolution of crude oil prices and evaluating how responsive they are to
underlying political and economic factors.

The Federal Reserve Economic Data (FRED) database, which aggregates,
verifies, and standardizes unprocessed market data, is the primary
source of the dataset, which shows a careful data collection method.
Daily WTI crude oil prices provide a real-time picture of market
circumstances and are based on trading activity on key exchanges.

Notwithstanding its advantages, the dataset's ability to accurately
reflect the complex variables affecting crude oil prices is limited.
Although they are not specifically represented, speculative trading
activities like the use of futures contracts are implicitly captured.
Likewise, unreported occurrences such as abrupt changes in geopolitics
or unreported production adjustments might affect pricing in ways that
are difficult for structured datasets to capture. Although the
election_phase variable is helpful for examining political impacts, it
oversimplifies complicated political processes into two main
categories—pre-election and post-election—and may miss more nuanced
market responses to changes in laws or regulations. These
representational issues draw attention to areas in which the dataset
might not accurately reflect the complex dynamics of the crude oil
market.

## 2.3 Outcome Variable

The outcome of interest in this analysis is the daily closing price of
Western Texas Intermediate (WTI) crude oil, denoted as wti_price. This
variable, measured in U.S. dollars, serves as a critical indicator of
the performance and valuation of crude oil in the global market. It
reflects real-time market dynamics, capturing the interplay between
supply and demand, geopolitical events, and broader economic trends.

Along with Brent Crude and Dubai/Oman, WTI crude oil is one of the three
main benchmarks for world oil prices. The Cushing, Oklahoma hub, a
crucial hub for the storage and delivery of oil in the United States,
has a direct impact on its pricing. Particularly for light, sweet crude
oil types, the WTI price affects global pricing dynamics and serves as a
benchmark for North American oil markets (EIA 2023).

Crude oil is a fungible commodity, therefore changes in supply and
demand have a significant impact on its price. While excess brought on
by increasing production, as occurs during OPEC's strategic output
expansions, tends to reduce prices, a spike in demand from major
economies like China and India frequently drives prices higher. WTI
prices are a trustworthy predictor of patterns in global energy use
since they nearly instantly reflect these equilibria (Hamilton 2009).

The data shows that WTI prices exhibit notable variability over time,
influenced by both short-term fluctuations and longer-term trends. For
instance, price levels may respond to daily market volatility, such as
changes in inventory levels or production decisions by OPEC, as well as
to more gradual shifts, such as those tied to election cycles or broader
economic growth. This underscores the importance of understanding the
factors driving WTI prices, as they have far-reaching implications for
energy markets, inflation, and economic stability.

By focusing on this outcome variable, this analysis seeks to uncover the
key predictors influencing crude oil prices, such as temporal trends
(date_numeric), election-related effects (election_phase), and
short-term changes in market performance (daily_pct_change and
price_change). The variability in this outcome highlights the complexity
of oil markets and the need for a robust statistical framework to
disentangle the effects of these interrelated factors.

## 2.4 Predictor Variables

Numerous factors that are thought to affect WTI crude oil prices
(wti_price) are captured by the predictors included in this analysis.
These variables offer a thorough framework for analyzing changes in the
price of oil by representing market, political, and temporal dynamics. A
thorough explanation of each predictor is provided below, along with
graphics showing their distributions and connections to the outcome
variable.

date_numeric represents the number of days since the start of the
dataset, serving as a measure of time. This variable captures the
long-term trend in WTI prices, reflecting gradual increases or decreases
over the observation period.

```{r}
# Load necessary libraries
library(tidyverse)

# File path to the simulated data
file_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load the dataset
analysis_data <- read_csv(file_path)

# Ensure date_numeric is numeric and sort by date
analysis_data <- analysis_data %>%
  mutate(
    date = as.Date(date), # Ensure 'date' is in Date format
    date_numeric = as.numeric(date - min(date, na.rm = TRUE)) # Calculate days since start of dataset
  )

# Create the scatterplot
scatter_plot <- ggplot(analysis_data, aes(x = date_numeric, y = wti_price)) +
  geom_point(alpha = 0.7) + # Add points
  geom_smooth(method = "lm", color = "blue", se = FALSE) + # Add a trend line
  labs(
    title = "Figure 1: Scatterplot of WTI Price Against Date Numeric",
    x = "Date Numeric (Days Since Start of Dataset)",
    y = "WTI Price (USD)"
  ) +
  theme_minimal()

# Save the plot
ggsave(
  filename = "/home/rstudio/finalpaper/figure1_scatterplot.png",
  plot = scatter_plot,
  width = 8,
  height = 6
)

# Display the plot
print(scatter_plot)

```

Figure 1. Scatterplot of WTI Prices Over Time with Linear Trend Line.

A systematic decline in crude oil prices over the research period is
suggested by a scatterplot of wti_price against date_numeric (Figure 1),
which displays a distinct downward trend over time. Understanding the
temporal dynamics of oil markets is possible thanks to this
characteristic.

The dataset is divided into two eras by the categorical variable
election_phase: "pre-election" (before to the 2024 U.S. Presidential
Election) and "post-election" (after the election). This variable is
used to look into how political events could affect WTI prices.

```{r}
# Load necessary libraries
library(tidyverse)

analysis_data <- read_csv(file_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )

# Ensure the election_phase variable is a factor
analysis_data <- analysis_data %>%
  mutate(
    election_phase = factor(election_phase, levels = c("pre-election", "post-election"))
  )

# Create the boxplot
boxplot_plot <- ggplot(analysis_data, aes(x = election_phase, y = wti_price, fill = election_phase)) +
  geom_boxplot(alpha = 0.7) + # Add boxplots
  scale_fill_manual(values = c("pre-election" = "red", "post-election" = "blue")) + # Custom colors
  labs(
    title = "Figure 2: Boxplot of WTI Price by Election Phase",
    x = "Election Phase",
    y = "WTI Price (USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Remove legend

# Save the plot
ggsave(
  filename = "/home/rstudio/finalpaper/figure2_boxplot.png",
  plot = boxplot_plot,
  width = 8,
  height = 6
)

# Display the plot
print(boxplot_plot)

```

Figure 2. Boxplot of WTI Prices Across Pre-Election and Post-Election
Phases.

A boxplot (Figure 2) compares wti_price distributions across the two
election phases. The pre-election phase shows consistently lower median
prices compared to the post-election phase, highlighting a possible
relationship between political cycles and market confidence.

daily_pct_change measures the percentage change in WTI price relative to
the previous day. This variable captures short-term market volatility,
reflecting rapid fluctuations influenced by supply-demand dynamics,
geopolitical events, or speculative trading.

```{r}
# Load necessary libraries
library(tidyverse)

# File path to the simulated data
file_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load the dataset
analysis_data <- read_csv(file_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )
# Create the histogram for daily_pct_change
histogram_plot <- ggplot(analysis_data, aes(x = daily_pct_change)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightblue", alpha = 0.7) + # Histogram settings
  labs(
    title = "Figure 3: Histogram of Daily Percentage Change in WTI Price",
    x = "Daily Percentage Change (%)",
    y = "Frequency"
  ) +
  theme_minimal()

# Display the plot
print(histogram_plot)

```

Figure 3. Histogram of Daily Percentage Changes in WTI Prices.

A histogram (Figure 3) shows the distribution of daily_pct_change,
centered around zero with a wide spread. This variability highlights the
often unpredictable nature of short-term oil price movements.

price_change represents the absolute dollar change in WTI price from one
day to the next. As an alternative measure of volatility, this variable
complements daily_pct_change by focusing on the magnitude of price
movements rather than their relative scale.

```{r}
# Load necessary libraries
library(tidyverse)


# File path to the simulated data
file_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"
analysis_data <- read_csv(file_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )

# Create the density plot for price_change
density_plot <- ggplot(analysis_data, aes(x = price_change)) +
  geom_density(fill = "purple", alpha = 0.7) + # Density plot settings
  labs(
    title = "Figure 4: Density Plot of Price Change in WTI Prices",
    x = "Price Change (USD)",
    y = "Density"
  ) +
  theme_minimal()

# Display the plot
print(density_plot)

```

Figure 4. Density Plot of Absolute Price Changes in WTI Prices.

The distribution of price_change is depicted in a density plot (Figure
4). The data reflects the episodic character of major market occurrences
with a fairly symmetric distribution around zero and sporadic
substantial movements.

The inclusion of each predictor is based on how well it explains
wti_price, but it's also important to consider how they interact. For
instance, election_phase may interact with temporal trends
(date_numeric) to enhance or reduce price movements at particular
periods, while daily_pct_change and price_change are anticipated to
represent distinct aspects of market volatility.

Together, these predictors provide a strong framework for comprehending
the intricate dynamics of WTI crude oil prices, which serves as the
basis for the statistical modeling used in this investigation. Appendix
A offers summary data tables and visualizations for these predictors for
additional study. The data shows a roughly symmetric distribution
centered around zero, with occasional large shifts, reflecting the
episodic nature of significant market events.

## 2.5 Associations between variables

Figure 5 illustrates the percentage of WTI price categories ("Low,"
"Medium," and "High") throughout the "pre-election" and "post-election"
phases, which sheds light on how price distributions change during these
times. The dataset counts the occurrences of each pricing category and
is classified by election_phase. To standardize the counts, proportions
are computed for every election phase. For clarity, each category is
given a different hue. To visualize the proportions, use
geom_bar(position = "fill") to generate a stacked proportional bar
chart.

```{r}
# Load necessary libraries
library(tidyverse)

# File path to the simulated data
file_path <- "/home/rstudio/finalpaper/data/00-simulated_data/simulated_data.csv"

# Categorize prices into low, medium, and high price ranges
analysis_data <- analysis_data %>%
  mutate(
    price_category = case_when(
      wti_price < 68 ~ "Low",
      wti_price >= 68 & wti_price < 70 ~ "Medium",
      wti_price >= 70 ~ "High"
    )
  )

# Create a proportion table for visualization
price_category_summary <- analysis_data %>%
  count(price_category, election_phase) %>%
  group_by(election_phase) %>%
  mutate(proportion = n / sum(n))

# Create the bar chart
price_category_plot <- ggplot(price_category_summary, aes(x = election_phase, y = proportion, fill = price_category)) +
  geom_bar(stat = "identity", position = "fill", alpha = 0.8) +
  scale_fill_manual(values = c("Low" = "red", "Medium" = "orange", "High" = "green")) +
  labs(
    title = "Proportion of Price Categories Across Election Phases",
    x = "Election Phase",
    y = "Proportion",
    fill = "Price Category"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# Save the plot
ggsave(
  filename = "/home/rstudio/finalpaper/price_category_plot.png",
  plot = price_category_plot,
  width = 8,
  height = 6
)

# Display the plot
print(price_category_plot)

```

Figure 5. Proportional Distribution of WTI Price Categories Across
Election Phases.

## 3 Model

# 3.1 Model set-up

To investigate the factors influencing WTI crude oil price fluctuations,
this study utilizes daily data spanning October 1 to November 18, 2024.
The dataset includes the WTI price, calculated daily percentage changes,
absolute price changes, and a numeric representation of the date. The
election phase variable is constructed to distinguish between
pre-election (October 1–November 4) and post-election (November 5–18)
periods.

A Bayesian linear regression model is employed with WTI price ((y_i)) as
the dependent variable and the predictors: date numeric ((x_1)),
election phase ((x_2)), daily percentage change ((x_3)), and absolute
price change ((x_4)):

A Bayesian linear regression model is employed with WTI price ((y_i)) as
the dependent variable and the predictors: date numeric ((x_1)),
election phase ((x_2)), daily percentage change ((x_3)), and absolute
price change ((x_4)):

\[ y_i \sim \mathcal{N}(\mu\_i, \sigma\^2) \]

\[ \mu\_i = \beta\_0 + \beta*1 x*{1i} + \beta*2 x*{2i} + \beta*3
x*{3i} + \beta*4 x*{4i} \]

\[ \beta\_0, \beta\_1, \beta\_2, \beta\_3, \beta\_4 \sim \mathcal{N}(0,
2.5) \]

\[ \sigma \sim \text{Exponential}(1) \]

Here, (\mathcal{N}) represents the normal distribution, and the priors
on the coefficients ((\beta)) are normal distributions centered at 0
with a standard deviation of 2.5. The variance parameter ((\sigma)) is
given an exponential prior with a rate of 1. The model leverages
informative priors to ensure robust estimation while accounting for
parameter uncertainty.

# 3.2 Model justification

The Bayesian framework offers a probabilistic approach to quantify
uncertainty in parameter estimates and assess the significance of
predictors in shaping WTI price fluctuations. This approach is
particularly suited to capturing the complexity of crude oil markets
influenced by intertwined economic, political, and temporal factors. By
incorporating election phase and temporal trends as predictors, the
model directly addresses the impact of political transitions and
sustained trends on price levels. Measures of short-term market
volatility (daily percentage and absolute price changes) are included to
evaluate their predictive contribution, allowing for a nuanced
exploration of both systematic and immediate price drivers.

```{r}
install.packages("DiagrammeR")
library(DiagrammeR)

library(DiagrammeR)

gr <- grViz("
digraph causal_model {
  rankdir=LR;
  
  ElectionPhase [label='Election Phase', shape=box];
  DateNumeric [label='Date Numeric', shape=box];
  DailyPctChange [label='Daily % Change', shape=box];
  AbsolutePriceChange [label='Absolute Price Change', shape=box];
  WTI_Price [label='WTI Price', shape=circle];
  
  ElectionPhase -> WTI_Price;
  DateNumeric -> WTI_Price;
  DailyPctChange -> WTI_Price;
  AbsolutePriceChange -> WTI_Price;
  ElectionPhase -> DateNumeric;
  ElectionPhase -> DailyPctChange;
}
")
```

Figure 6. Causal relationships between predictors (Election Phase, Date
Numeric, etc.) and the dependent variable (WTI Price).

## 4 Results

In this section, we visualized our data through graphs and tables as
well as present the results from our model.

# 4.1.1. Election phase and price dynamics

```{r}
# Load necessary libraries
library(tidyverse)
library(gt)

# File path to the simulated data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )


# Calculate summary statistics for WTI prices by election phase
summary_stats <- analysis_data %>%
  group_by(election_phase) %>%
  summarise(
    Mean = mean(wti_price, na.rm = TRUE),
    Median = median(wti_price, na.rm = TRUE),
    Range = paste0(
      round(min(wti_price, na.rm = TRUE), 2), " - ", round(max(wti_price, na.rm = TRUE), 2)
    )
  )

# Display the summary statistics in a formatted table
summary_stats %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics of WTI Prices by Election Phase",
    subtitle = "Mean, Median, and Range"
  ) %>%
  cols_label(
    election_phase = "Election Phase",
    Mean = "Mean Price",
    Median = "Median Price",
    Range = "Price Range"
  ) %>%
  fmt_number(
    columns = c(Mean, Median),
    decimals = 2
  )

```

Figure 7. Summary Statistics of WTI Prices by Election Phase: Mean,
Median, and Range.

Elections are often accompanied by increased market uncertainty,
particularly regarding future trade, energy policies, and geopolitical
relationships. These uncertainties can depress market confidence and
impact crude oil prices.

For example, concerns about potential shifts in U.S. energy policy,
including tariffs or sanctions, could drive anticipatory price
adjustments in the market. As election results define the political
environment and give markets time to stabilize and adapt to expected
policies, the post-election period typically signifies a period of
decreased uncertainty. Restoring investor confidence may result in
higher prices if political parties or administrations place a strong
emphasis on domestic energy independence, whether through overt
messaging or policy signals.

A strong dollar negatively impacts oil prices since oil is traded in
dollars, making it more expensive for foreign buyers. According to
analysts, the initial concerns about oversupply caused by Trump's
initiatives were exaggerated. While John Kilduff (Again Capital)
observes that the market has adjusted following initial over-reactions,
Phil Flynn (Price Futures Group) sees additional upside possibilities in
the near term.

# 4.1.2. Daily Volatility and Price Changes

Oil prices fell more than 1% on Wednesday, November 6th, after Donald
Trump was elected president of the United States. The main cause of this
decline was the strengthening of the US dollar, which hit its highest
level since March 2023.

Besides the WTI, Brent crude oil futures were also down \$1.00, 0r 1.32
percent, at \$74.53 per barrel by 1040 GMT, while U.S. West Texas
Intermediate (WTI) crude fell 93 cents, or 1.29 percent, to \$71 per
barrel.

UBS analyst Giovanni Staunovo said, "A Trump presidency has a bearish
spin. Tariffs would be negative for economic growth and oil demand
growth. However, Trump could renew sanctions on Iran and Venezuela,
removing barrels from the market, which would be bullish.” Iran exports
about 1.3 million barrels per day.

```{r}
# Load necessary libraries
library(tidyverse)

# File path to the simulated data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load the data
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),  # Add comma here
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )

# Create a histogram of daily percentage changes by election phase
histogram_plot <- analysis_data %>%
  ggplot(aes(x = daily_pct_change, fill = election_phase)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 30) +
  scale_fill_manual(values = c("Pre-Election" = "blue", "Post-Election" = "red")) +
  labs(
    title = "Histogram of Daily Percentage Changes",
    x = "Daily Percentage Change (%)",
    y = "Count",
    fill = "Election Phase"
  ) +
  theme_minimal()

# Create a density plot of daily percentage changes by election phase
density_plot <- analysis_data %>%
  ggplot(aes(x = daily_pct_change, fill = election_phase)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Pre-Election" = "blue", "Post-Election" = "red")) +
  labs(
    title = "Density Plot of Daily Percentage Changes",
    x = "Daily Percentage Change (%)",
    y = "Density",
    fill = "Election Phase"
  ) +
  theme_minimal()

# Display the plots
print(histogram_plot)
print(density_plot)


```

Figure 8 and 9. Comparison of the Distribution of Daily Percentage
Changes in WTI Prices Across Election Phases and Density Plot
Highlighting Daily Percentage Changes in WTI Prices During Pre- and
Post-Election Phases.

# 4.1.3 Temporal trends

```{r}
# Load necessary libraries
library(tidyverse)

# File path to the simulated data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load the data
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    date_numeric = as.numeric(date - min(date, na.rm = TRUE))  # Convert dates to numeric for plotting
  )

# Create a scatter plot with a regression line
scatter_plot <- analysis_data %>%
  ggplot(aes(x = date_numeric, y = wti_price)) +
  geom_point(alpha = 0.6, color = "blue", size = 2) +  # Scatter points
  geom_smooth(method = "lm", color = "red", se = TRUE, linetype = "dashed") +  # Regression line
  labs(
    title = "Scatter Plot of WTI Prices Over Time",
    x = "Numeric Date (Days since start)",
    y = "WTI Price ($)"
  ) +
  theme_minimal()

# Save the plot as a PNG file
ggsave("scatter_plot_wti_vs_date_numeric.png", scatter_plot, width = 8, height = 6)

# Display the plot
print(scatter_plot)

```

Figure 10. Scatter Plot of WTI Prices Over Time with a Linear Regression
Line.

The Middle East conflict may cause supply disruptions for oil, which
would drive up prices. Trump's backing of Israel could make things more
unstable in the region. Bullish conditions could result from the removal
of large output (such as Iran's 3.2 million barrels per day) from the
market if Trump decides to prolong the sanctions.

Analysts point out that Iran has mastered the art of dodging sanctions,
which may lessen its effect. With enduring problems including refinery
margins, declining demand, and inefficiencies impacting market dynamics,
OPEC+ continues to play a crucial role in regulating world supply.
Beyond geopolitical uncertainty, analysts like Rystad Energy's Mukesh
Sahdev stress that fundamental market trends like refinery margins,
OPEC+ control, and trade inefficiencies are crucial in determining the
future paths of oil prices.

# 4.1.4. Model Coefficients and Predictive Insights

The article "Understanding and interpreting confidence and credible
intervals around effect estimates" by Hespanhol et al. (2019) provides
insights into the interpretation of credible intervals within Bayesian
statistics. A credible interval represents a range within which an
unobserved parameter value lies with a certain probability, given the
observed data and prior information. This probabilistic interpretation
directly quantifies uncertainty about the parameter.

```{r}
# Load necessary libraries
library(tidyverse)
library(rstanarm)
library(bayesplot)

# File path to the data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load and preprocess the data
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    date_numeric = as.numeric(date - min(date, na.rm = TRUE)),  # Convert dates to numeric for plotting
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = (price_change / lag(wti_price)) * 100  # Calculate daily percentage change
  )

# Fit a Bayesian regression model
wti_price_model <- stan_glm(
  formula = wti_price ~ date_numeric + election_phase + daily_pct_change + price_change,
  data = analysis_data,
  family = gaussian(),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_aux = exponential(rate = 1, autoscale = TRUE),
  seed = 853
)

# Extract posterior samples for key coefficients
posterior_samples <- as.matrix(wti_price_model)

# Generate MCMC areas plot for key coefficients
mcmc_areas_plot <- mcmc_areas(
  posterior_samples,
  pars = c("(Intercept)", "date_numeric", "election_phasePre-Election", "daily_pct_change", "price_change"),
  prob = 0.8  # 80% credible intervals
) +
  labs(
    title = "Posterior Distributions of Key Coefficients",
    x = "Posterior Density",
    y = "Coefficients"
  ) +
  theme_minimal()

# Save the plot as a PNG file
ggsave("mcmc_areas_plot.png", mcmc_areas_plot, width = 8, height = 6)

# Display the plot
print(mcmc_areas_plot)

```

Figure 11. Posterior Distributions of Key Coefficients with 80% Credible
Intervals

```{r}
# Load necessary libraries
library(tidyverse)
library(rstanarm)
library(broom.mixed) # Use broom.mixed for rstanarm models
library(knitr)
library(kableExtra) # For formatted tables in Quarto

# File path to the data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load and preprocess the data
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    date_numeric = as.numeric(date - min(date, na.rm = TRUE)),  # Convert dates to numeric for plotting
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = ifelse(!is.na(lag(wti_price)) & lag(wti_price) != 0,
                              (price_change / lag(wti_price)) * 100, NA)  # Calculate daily percentage change
  )

# Fit a Bayesian regression model
wti_price_model <- stan_glm(
  formula = wti_price ~ date_numeric + election_phase + daily_pct_change + price_change,
  data = analysis_data,
  family = gaussian(),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_aux = exponential(rate = 1, autoscale = TRUE),
  seed = 853
)

# Extract and tidy the model results using broom.mixed
results_df <- broom.mixed::tidy(wti_price_model, conf.int = TRUE, conf.level = 0.95)

# Create a formatted table for Quarto
results_table <- results_df %>%
  kable(
    format = "html",
    caption = "Bayesian Regression Results",
    col.names = c("Term", "Estimate", "Std. Error", "Lower 95% CI", "Upper 95% CI")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

# Save the table as an HTML file for Quarto
save_kable(results_table, file = "bayesian_regression_results.html")

# Print the table
print(results_table)


```

Figure 12. Full set of outputted results

# 4.1.5. Posterior Predictive Checks

```{r}
# Load necessary libraries
library(tidyverse)
library(rstanarm)
library(bayesplot)

# File path to the data
data_path <- "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Load and preprocess the data
analysis_data <- read_csv(data_path) %>%
  mutate(
    date = as.Date(date),  # Ensure the date column is in Date format
    date_numeric = as.numeric(date - min(date, na.rm = TRUE)),  # Convert date to numeric
    election_phase = ifelse(date < as.Date("2024-11-05"), "Pre-Election", "Post-Election"),
    election_phase = as.factor(election_phase),  # Convert to factor
    daily_pct_change = ifelse(!is.na(lag(wti_price)) & lag(wti_price) != 0,
                              (price_change / lag(wti_price)) * 100, NA)  # Calculate daily percentage change
  )

# Check for missing columns
required_columns <- c("wti_price", "price_change", "daily_pct_change")
missing_columns <- setdiff(required_columns, colnames(analysis_data))
if (length(missing_columns) > 0) {
  stop(paste("Missing columns in dataset:", paste(missing_columns, collapse = ", ")))
}

# Fit a Bayesian regression model
wti_price_model <- stan_glm(
  formula = wti_price ~ date_numeric + election_phase + daily_pct_change + price_change,
  data = analysis_data,
  family = gaussian(),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_aux = exponential(rate = 1, autoscale = TRUE),
  seed = 853
)

# Generate posterior predictive checks
pp_check_plot <- pp_check(wti_price_model, type = "dens_overlay") +
  labs(
    title = "Posterior Predictive Check: Observed vs. Predicted WTI Prices",
    x = "WTI Price",
    y = "Density"
  ) +
  theme_minimal()

# Save the plot as a PNG file
ggsave("posterior_predictive_check.png", pp_check_plot, width = 8, height = 6)

# Display the plot
print(pp_check_plot)
```

Figure 13. Graph of posterior predictive checks comparing observed and
predicted WTI prices.

## 4.2 Model Results

```{r}
# Load necessary libraries
library(bayesplot)
library(tidyverse)

# This script assumes `wti_price_model` has been fitted using the dataset located at:
# "/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv"

# Extract posterior samples
posterior_samples <- as.matrix(wti_price_model)

# Check available parameter names in posterior samples
available_pars <- colnames(posterior_samples)
print("Available Parameters:")
print(available_pars)

# Plot credible intervals (ensure correct parameter names)
coef_plot <- mcmc_areas(
  posterior_samples,
  pars = c("(Intercept)", "date_numeric", "election_phasePre-Election", 
           "daily_pct_change", "price_change"),  # Correct parameter names based on output
  prob = 0.90  # 90% credible intervals
) +
  labs(
    title = "Posterior Distributions of Coefficients",
    subtitle = "90% Credible Intervals",
    x = "Posterior Estimate",
    y = "Coefficients"
  ) +
  theme_minimal()

# Save and display plot
ggsave("coefficient_plot.png", coef_plot, width = 8, height = 6)
print(coef_plot)


```

Figure 15. Coefficient Plot with Credible Intervals.

When all predictor variables are at their baseline or reference levels,
the intercept (B0) shows the average expected price of West Texas
Intermediate (WTI) crude oil. For example, the baseline levels match the
reference categories or mean values of the predictors, depending on the
kind of variable, if the model incorporates predictors such as market
demand, production levels, and geopolitical stability.

In this instance, the intercept value of \$77.9 indicates that when the
predictors are at these baseline levels, the expected WTI price is
approximately \$77.9. This serves as a starting point for the model’s
predictions, and deviations from this value are explained by the other
predictors in the model.

The degree to which the observed values differ from the intercept's
predicted values is indicated by the Mean Absolute Deviation (MAD). A
low MAD, as in this case, reflects minimal variability, indicating that
the model is suitably stable in estimating this baseline price. This
stability is crucial for building confidence in the model's overall
predictive power, as a highly variable intercept might signal issues
like model misspecification or insufficient data quality.

An average daily decline of \$0.2 in WTI prices is indicated by the
predicted coefficient for the numeric representation of dates (B1),
which is -0.2. This negative trend indicates a general temporal fall in
prices across the observation period, which may have been caused by more
general economic issues and increased market hesitancy as a result of
Trump's frequent remarks about speculative tariffs. The gradual decline
in oil prices over time seems to be a result of these factors.

The coefficient for daily percentage changes (B3) is small, estimated at
0.5, with a relatively large mean absolute deviation (MAD) of 1.5. This
suggests that while daily market volatility exists, its predictive power
for overall price levels is limited. The wide variability in this
predictor indicates that short-term fluctuations may not have a
consistent or significant impact on WTI prices.

With large credible intervals, the coefficient for daily absolute price
changes (B4) is -0.3, which is close to zero. Individual daily price
changes may not have a substantial or consistent impact on overall price
levels, according to this weak correlation. The large gaps emphasize
even more how difficult it is to assign this variable any meaningful
significance.

The model diagnostics, including measures like R2 and the mean posterior
predictive distribution (mean_PPD), indicate excellent convergence and
stability, with all R2 values equal to 1. Posterior predictive checks
confirm strong alignment between observed and predicted price
distributions, affirming the model's ability to capture key patterns and
trends in the data.

## 5 Limitations and Next Steps

# 5.1 Limitations

The dataset comprises only 32 observations, limiting its capacity to
capture long-term trends and seasonal patterns in WTI crude oil prices.
Such a brief timeframe may result in biased estimates, especially for
predictors like date_numeric, which could indicate a gradual decline
over time.

To uncover underlying patterns, time series analysis breaks down data
into trend, seasonality, and residual components. However, this
decomposition is hampered by an inadequate dataset, making it difficult
to precisely identify and model these components. As a result, the model
can miss important trends, producing inaccurate predictions.

According to study published in the Journal of Time Series Econometrics,
sufficient data length is crucial for efficient time series
decomposition. Ollech (2021), for example, addresses the difficulties of
correcting daily time series data for seasonal effects in "Seasonal
Adjustment of Daily Time Series," emphasizing that incomplete data can
make it difficult to accurately estimate seasonal and trend components.

Adding interaction terms could improve the model's interpretability and
forecast performance by better reflecting the intricacy of the
underlying processes. Election_phase and daily_pct_change, for example,
may not have a strictly additive connection. An election's phase may
interact with other elements, such geopolitical events or economic data,
to increase or decrease market volatility. For instance, increased
political unpredictability during crucial election periods may cause
investor concern to cause inflated daily percentage changes in the
market. The model runs the danger of oversimplifying these dynamics and
missing how the interaction of various variables influences market
behavior if such interaction effects are ignored.

The model ignores external factors like global oil demand, OPEC+
decisions, and macroeconomic indices like GDP or inflation in favour of
concentrating only on a small number of predictors. Since these factors
have a big impact on oil prices, leaving them out could confuse the
links that are shown. A more thorough knowledge of price changes may be
obtained by including these factors in the model. Ignoring them could
result in omitted variable bias, which would skew the estimated effects
of the predictors that are included.

The posterior predictive distribution (mean_PPD) of the model only
roughly fits the observed data, despite the fact that R2 values show
high convergence. This raises the possibility of missing explanatory
variables or problems with model formulation.

# 5.2 Next Seps

Expand the dataset to include observations spanning multiple election
cycles and periods of economic volatility. The many temporal and
political factors affecting WTI pricing would be more accurately
captured by wider temporal coverage. The investigation of trends over
long periods of time may also be enhanced by the inclusion of time
series data with fixed effects for time and country.

Include interaction terms to examine how political events affect market
reactions, such as between election_phase and daily_pct_change. The
model's explanatory strength and predictive accuracy may be enhanced by
such interaction effects, which could reveal intricate interactions
between predictors. One could also improve the model by adding external
factors like the demand for crude oil around the world, OPEC+ production
choices, and macroeconomic metrics like GDP growth or inflation. These
elements would offer crucial background information, allowing for a more
thorough examination of price swings in relation to larger market and
geopolitical issues.

Uncertainty in predictors with broad credible intervals could be better
taken into account by modifying the priors or adding hierarchical
structures. These improvements may improve the accuracy of predictions
for short-term market volatility forecasters. To thoroughly assess model
fit and pinpoint areas in need of development, carry out extra posterior
predictive assessments, such as residual diagnostics.

Run simulations under fictitious conditions, such as increased sanctions
on Iran or significant OPEC+ output cuts. Actionable insights into the
possible effects of particular geopolitical or economic events on WTI
prices could be obtained from these simulations.

# Appendix {.unnumbered}

## A.1 Sketches

Data and graph sketches can be found in this report’s GitHub Repository.

## A.2 R Scripts

Scripts with code used for simulation, testing, downloading, data
cleaning, modelling, replicating and exploratory data analysis can be
found in this report’s GitHub Repository. The R programming language (R
Core Team 2023) as well as the tidyverse package (Wickham et al. 2019)
were used to test the real and simulated data.

## A.3 Additional Plots

```{r}
#### Load data ####
analysis_data <- read_csv("/home/rstudio/finalpaper/data/02-analysis_data/analysis_data.csv")

#### 1. WTI Prices Over Time ####
wti_time_plot <- ggplot(analysis_data, aes(x = date, y = wti_price)) +
  geom_line() +
  labs(
    title = "WTI Crude Oil Prices Over Time",
    x = "Date",
    y = "Price (USD per barrel)"
  ) +
  theme_minimal()

print(wti_time_plot)

```

Figure 16. WTI Prices Over Time.

```{r}
#### 2. Histogram of WTI Prices ####
wti_histogram <- ggplot(analysis_data, aes(x = wti_price)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of WTI Crude Oil Prices",
    x = "Price (USD per barrel)",
    y = "Frequency"
  ) +
  theme_minimal()

print(wti_histogram)

```

Figure 17. Histogram of WTI Prices.

```{r}
#### 3. Boxplot to Detect Outliers ####
wti_boxplot <- ggplot(analysis_data, aes(y = wti_price)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Boxplot of WTI Crude Oil Prices",
    y = "Price (USD per barrel)"
  ) +
  theme_minimal()

print(wti_boxplot)

```

Figure 18. Boxplot to Detect Outliers.

```{r}
#### 4. Monthly Average Prices ####
analysis_data_monthly <- analysis_data %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(avg_price = mean(wti_price, na.rm = TRUE))

monthly_avg_plot <- ggplot(analysis_data_monthly, aes(x = month, y = avg_price)) +
  geom_line(color = "blue") +
  labs(
    title = "Monthly Average WTI Crude Oil Prices",
    x = "Month",
    y = "Average Price (USD per barrel)"
  ) +
  theme_minimal()

print(monthly_avg_plot)

```

Figure 19. Monthly Average Prices.

```{r}
#### 6. Daily Percentage Change in Prices ####
analysis_data <- analysis_data %>%
  mutate(daily_pct_change = 100 * (wti_price - lag(wti_price)) / lag(wti_price))

daily_pct_change_plot <- ggplot(analysis_data, aes(x = date, y = daily_pct_change)) +
  geom_line(color = "red", alpha = 0.7) +
  labs(
    title = "Daily Percentage Change in WTI Crude Oil Prices",
    x = "Date",
    y = "Percentage Change (%)"
  ) +
  theme_minimal()

print(daily_pct_change_plot)

```

Figure 20. Daily Percentage Change in Prices.

```{r}
#### 7. Boxplot by Election Phase ####
election_date <- as.Date("2024-11-05")
analysis_data <- analysis_data %>%
  mutate(
    election_phase = ifelse(date < election_date, "pre-election", "post-election"),
    election_phase = as.factor(election_phase)
  )

election_phase_boxplot <- ggplot(analysis_data, aes(x = election_phase, y = wti_price, fill = election_phase)) +
  geom_boxplot() +
  labs(
    title = "WTI Crude Oil Prices by Election Phase",
    x = "Election Phase",
    y = "Price (USD per barrel)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(election_phase_boxplot)

```

Figure 21. Boxplot by Election Phase.

## A.4 Posterior Predictive Check

\section{Posterior Predictive Check}

Posterior predictive checks are an essential diagnostic tool for
assessing how well the fitted Bayesian model aligns with the observed
data. Figure\~\ref{fig:ppcheck} provides a visual comparison of the
observed and predicted distributions, helping to evaluate the model's
ability to capture the variability and central tendencies of the data.

```{=tex}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\textwidth]{path_to_ppcheck_plot.png}
    \caption{Posterior predictive check comparing observed and predicted distributions of WTI prices. The shaded region represents the variability captured by the model.}
    \label{fig:ppcheck}
\end{figure}
```
In addition to checking the predictive fit, we also compare the
posterior distributions of the key model parameters with their priors to
assess how much information the data contributed to the posterior
estimates. Figure\~\ref{fig:priorvsposterior} demonstrates this
comparison, where significant shifts between the prior and posterior
indicate strong updates driven by the observed data.

```{=tex}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\textwidth]{path_to_prior_vs_posterior_plot.png}
    \caption{Comparison of posterior and prior distributions for key model parameters. The posterior distributions are updated based on observed data, reflecting evidence-driven inferences.}
    \label{fig:priorvsposterior}
\end{figure}
```
Overall, these diagnostic checks confirm that the model is
well-calibrated, capturing the observed data's essential features while
allowing for meaningful inference about key predictors.

## A.5 Diagnostics

\section{Diagnostics}

To ensure the validity of the Bayesian model, we assess convergence and
sampling quality using diagnostic plots. Figure\~\ref{fig:diagnostics}
presents two key diagnostics: a trace plot and an (\hat{R}) plot. These
plots help verify whether the Markov Chain Monte Carlo (MCMC) sampling
process has converged and whether the posterior estimates are reliable.

```{=tex}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.45\textwidth]{path_to_trace_plot.png}
    \includegraphics[width=0.45\textwidth]{path_to_rhat_plot.png}
    \caption{Checking the convergence of the MCMC algorithm. (Left) The trace plot shows the sampling paths for the model's key parameters across iterations. (Right) The \(\hat{R}\) plot indicates convergence metrics for each parameter.}
    \label{fig:diagnostics}
\end{figure}
```
\subsection{Trace Plot}

The trace plot in Figure\~\ref{fig:diagnostics} (left) illustrates the
sampling paths for key parameters across the MCMC iterations. The
well-mixed and stationary paths indicate that the sampler has explored
the posterior distribution efficiently. This suggests that the chains
have converged and the samples are representative of the target
posterior distribution.

\subsection{\(\hat{R}\) Plot}

The (\hat{R}) plot in Figure\~\ref{fig:diagnostics} (right) shows the
potential scale reduction factor ((\hat{R})) for each parameter. All
(\hat{R}) values are close to 1, indicating that the chains have
converged and there are no issues with mixing or insufficient sampling.
These results confirm the stability and reliability of the posterior
estimates.

\newpage

# References

@book{alexander2023telling, title = {Telling Stories with Data}, author
= {Rohan Alexander}, year = {2023}, publisher = {CRC Press}, address =
{Boca Raton}, url = {https://tellingstorieswithdata.com/} }

@article{arelbundock2022modelsummary, title = {Modelsummary: Data and
Model Summaries in R}, author = {Vincent Arel-Bundock}, journal =
{Journal of Statistical Software}, volume = {103}, number = {1}, pages =
{1--23}, year = {2022}, doi = {10.18637/jss.v103.i01} }

@misc{journal_time_series, title = {Journal of Time Series
Econometrics}, publisher = {Berkeley Electronic Press}, address =
{Berkeley, CA}, year = {2008} }

@article{hespanhol2018confidence, title = {Understanding and
Interpreting Confidence and Credible Intervals around Effect Estimates},
author = {Luiz Hespanhol and Caio Sain Vallio and Lucíola Menezes Costa
and Bruno T. Saragiotto}, journal = {Brazilian Journal of Physical
Therapy}, volume = {23}, number = {4}, pages = {290--301}, year =
{2018}, doi = {10.1016/j.bjpt.2018.12.006} }

@misc{eia2024wti, title = {Crude Oil Prices: West Texas Intermediate
(WTI) - Cushing, Oklahoma \[DCOILWTICO\]}, author = {U.S. Energy
Information Administration}, note = {Retrieved from FRED, Federal
Reserve Bank of St. Louis. Accessed December 3, 2024}, url =
{https://fred.stlouisfed.org/series/DCOILWTICO} }

@article{sharma2024trump, title = {Donald Trump’s Presidency Will Likely
Boost U.S. Oil Output in 2025}, author = {Gaurav Sharma}, journal =
{Forbes}, year = {2024}, note = {November 6, 2024}, url =
{https://www.forbes.com/sites/gauravsharma/2024/11/06/donald-trumps-presidency-will-likely-boost-us-oil-output-in-2025/}
}

@misc{trump2024tariffs, title = {Energy Experts Think Donald Trump Will
Make Tariff Exemptions for Canadian Oil}, author = {Donald Trump},
journal = {CTVNews}, year = {2024}, note = {November 14, 2024}, url =
{https://www.ctvnews.ca/politics/energy-experts-think-donald-trump-will-make-tariff-exemptions-for-canadian-oil-1.7109771}
}

@misc{cnbc2024oil, title = {Oil Markets' Future Still Uncertain under
Donald Trump after Election Win}, author = {CNBC}, year = {2024}, note =
{November 6, 2024}, url =
{https://www.cnbc.com/2024/11/06/oil-markets-future-still-uncertain-under-donald-trump-after-election-win.html}
}

@misc{lefebvre2024oil, title = {Will Trump’s Win Be a Pyrrhic Victory
for the Oil Industry?}, author = {Ben Lefebvre}, journal = {Politico},
year = {2024}, note = {November 7, 2024}, url =
{https://www.politico.com/news/2024/11/07/trump-win-pyrrhic-victory-for-oil-industry-00187559}
}

@misc{king2024tariffs, title = {Republicans Praise Trump Tariff Threat
as Smart ‘Negotiating Strategy’ as He Looks to Tackle Fentanyl, Border
Crises}, author = {Ryan King and Josh Christenson}, journal = {New York
Post}, year = {2024}, note = {November 27, 2024}, url =
{https://nypost.com/2024/11/26/us-news/republicans-praise-trump-tariff-threat-as-smart-negotiating-strategy/?utm_source=chatgpt.com}
}

@misc{bloomberg2024energy, title = {From Oil to EVs, Here’s What a Trump
Victory Means for Energy}, author = {Bloomberg News}, journal =
{Financial Post}, year = {2024}, note = {November 6, 2024}, url =
{https://financialpost.com/commodities/energy/from-oil-to-evs-heres-what-a-trump-victory-means-for-energy}
}

@misc{guardian2024fundraising, title = {Trump's Oil and Gas Fundraising
Surges Ahead of Election}, journal = {The Guardian}, year = {2024}, note
= {October 18, 2024}, url =
{https://www.theguardian.com/us-news/2024/oct/18/election-trump-oil-gas-fundraising}
}

@article{baumeister2013oilshocks, title = {Time-varying effects of oil
supply shocks on the US economy}, author = {Baumeister, C. and Peersman,
G.}, journal = {American Economic Journal: Macroeconomics}, volume =
{5}, number = {4}, pages = {1--28}, year = {2013} }

@article{hamilton2009oilshock, title = {Causes and consequences of the
oil shock of 2007-08}, author = {Hamilton, J. D.}, journal = {Brookings
Papers on Economic Activity}, volume = {2009}, number = {1}, pages =
{215--261}, year = {2009} }

@inbook{hamilton2011resources, title = {Oil prices, exhaustible
resources, and economic growth}, author = {Hamilton, J. D.}, booktitle =
{Handbook of Energy and Climate Change}, volume = {1}, pages = {29--56},
year = {2011} }

@article{kilian2014inventories, title = {The role of inventories and
speculative trading in the global market for crude oil}, author =
{Kilian, L. and Murphy, D. P.}, journal = {Journal of Applied
Econometrics}, volume = {29}, number = {3}, pages = {454--478}, year =
{2014} }
